A Switching Planner for Combined Task and Observation Planning

### 2011-02-05 subarchitectures/planner.sa/trunk/src/latex/AAAI11/dtpddl.tex
Only one extra column to remove now. Any ideas people?!

\vspace{-1ex}
\small
\begin{tabtt}
(\=:init (= (is-in DORA) kitchen) \+ \\
       (probabilistic \=.8 (= (is-in box) office)  \\
		      \>.2 (= (is-in box) kitchen)) \\
       (probabilistic .3 (= (is-in cup) office)  \\
		      \>.7 (= (is-in cup) kitchen))) \\
\end{tabtt}
\normalsize

@@

\small
\begin{tabular}{cccc}
\hline
Probability & (is-in DORA)  & (is-in box)  & (is-in cup) \\
\hline
.24 & kitchen & office & office \\
.06 & kitchen & kitchen & office \\
.56 & kitchen & office & kitchen \\
.14 & kitchen & kitchen & kitchen \\
\hline
\end{tabular}
\normalsize

@@

>>>>

\vspace{-1ex}
\small
\begin{tabtt}
(\=:init (= (is-in R2D2) kitchen) \+ \\
       (probabilistic \=.8 (= (is-in box) office)  \\
		      \>.2 (= (is-in box) kitchen)) \\
       (probabilistic .3 (= (is-in cup) office)  \\
		      \>.7 (= (is-in cup) kitchen))) \\
\end{tabtt}
\normalsize

@@

\small
\begin{tabular}{cccc}
\hline
Probability & (is-in R2D2)  & (is-in box)  & (is-in cup) \\
\hline
.24 & kitchen & office & office \\
.06 & kitchen & kitchen & office \\
.56 & kitchen & office & kitchen \\
.14 & kitchen & kitchen & kitchen \\
\hline
\end{tabular}
\normalsize

@@

If a dual is executed, then the sequential planner must
replan with $\bstate_0$ equal to the underlying
belief-state. Otherwise, if $\actions.\poss(\switchAction)$ is
executed, then the current sequential plan is executed until further
sensing is scheduled, or to completion.

% \small
% \begin{tabtt}
% (\=:init (= (is-in DORA) kitchen) \+ \\
%        (probabilistic \=.8 (= (is-in box) office)  \\
% 		      \>.2 (= (is-in box) kitchen))) \\
% \end{tabtt}
% \normalsize

% \noindent and therefore, the belief-state $\bstate_0$ is:

% \small
% \begin{tabular}{cccc}
% \hline
% Probability & (is-in DORA)  & (is-in box)  & (is-in cup) \\
% \hline
% .8 & kitchen & office & \# \\
% .2 & kitchen & kitchen & \# \\
% \hline
% \end{tabular}
% \normalsize

 % With
% regards to abstraction, the contingent and serial sessions apply the
% same applicability conditions ---i.e., that each effect and condition
% $\prop$ is specified, $\prop\in\state$--- to actions from the DTPDDL
% problem description.

### 2011-02-05 subarchitectures/planner.sa/trunk/src/latex/AAAI11/switching.tex
Only one extra column to remove now. Any ideas people?!

\subsection{Sequential Planning in POMDPs}

A sequential session uses a classical planner to compute a trace. The
latter encapsulates assumptions about: (1) the true underlying state,
(2) how execution will progress, and (3) the possibility of the agent
eventually holding strong beliefs about the truth values of specific
state propositions. Here we describe the deterministic planning
problem, derived from the DTPDDL model, that admits plans which
correspond to traces.

In a deterministic-action POMDP all the uncertainty in state and
action is expressed in the $(\pp{:init})$ declaration. Our approach
uses the structure of that, as it occurs in the problem description,
to define a set of state-assumptions available to sequential planning.
%%
Writing \#\ if the value of a proposition is unspecified, for
$\pp{DORA}$ we have the following assumptions:

\small
\begin{tabular}{cccc}
\hline
Probability & (is-in DORA)  & (is-in box)  & (is-in cup) \\
\hline
.24 & kitchen & office & office \\
.06 & kitchen & kitchen & office \\
.56 & kitchen & office & kitchen \\
.14 & kitchen & kitchen & kitchen \\
.7 & kitchen & \# &  kitchen\\
.3 & kitchen & \# & office \\
.8 & kitchen & office & \# \\
.2 & kitchen & kitchen & \# \\
1.0 & kitchen & \# & \# \\
\hline
\end{tabular}
\normalsize

\noindent An assumption corresponds to a {\em
relaxed} visitation of the root term of $(\pp{:init})$. In this
relaxed case, a conjunctive term is visited iff its atomic subterms
are visited, and zero or more of its immediate probabilistic subterms
are visited. The starting state, $\state_0$, for sequential planning
is an {\em abstract} state that captures the unique assumption which
has probability $1$. In $\pp{DORA}$, that is:

\[
\begin{array}{l}
\state_0 \equiv \{(=(\pp{is-in}~\pp{DORA})~\pp{kitchen}),\\
\;\;(=(\pp{is-in}~\pp{box})~\#), (=(\pp{is-in}~\pp{cup})~\#)\}.
\end{array}
\]

In order to make assumptions available to the sequential planner, we
add one {\em assumptive} action $\assumptiveS{i}$ to the problem for
each element, $\prob_i (T_i)$, of each probabilistic term from
$(\pp{:init})$. The physics of these actions is as
follows: $\assumptiveS{i}$ can be executed if no $\assumptiveS{j}$,
$j \neq i$, has been executed from the same probabilistic term, and,
either $(\pp{probabilistic}~..\prob_i~(T_i)..)$ is in the root
conjunct, or it occurs in $T_k$ for some executed
$\assumptiveS{k}$. Also, an assumption about a proposition
$\prop$ can only be made once, and must be sequenced before any POMDP
actions that have $\prop$ in their precondition or effect.
%%
In other words, an assumptive action
must be applied to make a truth assignment to \prop, and then that
truth assignment can be modified by the DTPDDL domain operators.
%%
%%
Executing $\assumptiveS{i}$ in $\state$ effects a transition to a
successor state $\state^{T_i}$ with probability $\prob_i$, and
$\state^\bot$ with probability $1 - \prob_i$. Here, $\state^{T_i}$ is
the union of $\state$ with atomic terms from $T_i$ with the
proposition denoting the undefined fact is deleted. For example, if
the action adds $(=(\pp{is-in}~\pp{box})~\pp{office})$, then it must
delete $(=(\pp{is-in}~\pp{box})~\#)$. State $\state^\bot$ is an added
sink, where $\forall\action$ $\reward(\state^\bot, \action) = 0$ and
$\transProb(\state^\bot, \action, \state^\bot) = 1$.
%%
Lastly, we have that if $\action$ has a precondition that is not true
with probability $\theta$, a given threshold, in the underlying belief
$\bstate$, then immediately before it is executed blindly the system
switches to a contingent session (described below).

%% For sequential planning the operators from the POMDP model are
%% available with semantics that accommodate the above abstraction as
%% follows: In the deterministic model a proposition $\prop$ can be
%% thought to have a ternary interpretation at a state \state, as either
%% {\em true}, written $\prop\in\state$, {\em false}, written
%% $\prop\not\in\state$, or {\em unspecified}, with a slight abuse of
%% notation written $\prop\#\in\state$. For example, in $\pp{DORA}$
%% $(=(\pp{is-in}~\pp{box})\pp{office})\#\in\state_0$. For POMDP actions
%% $\action\in\actions$ in the deterministic model, if
%% $\prop \in \poss(\action)$, or if $\prop$ is the subject of a positive
%% or negative effect of $\action$, then $\action$ is not applicable
%% in \state\ if $\prop\#\in\state$.

%% Addressing now the switching semantics of action execution. I
%% %% In detail, we halt the sequential
%% session at \action\ if $\exists
%% \state$ s.t. $\bstate(\state)>0$, 
%% $\prop\in\poss(\action)$ , and $\prop\not\in\state$. According to the
%% semantics of action execution there must be a $\state'$ s.t.
%% $\bstate(\state')>0$ and $\prop\in\state'$. Otherwise, \action\ must
%% not have been scheduled at \bstate.

%% then $\reward(\state,\action)=10$, and
%% otherwise if $\prop\not\in\state'$ then
%% $\reward(\state',\action)=-10$.

We now describe the optimisation criteria given to the classical
planner. Where $\prob_i$ is the probability that the $i^{th}$
sequenced action $\action_i$ does not transition to $\state^\bot$, we
define the value of a trace
$\state_0, \action_0, \state_1, \action_1,.., \state_N$ from the
deterministic model to equal:

\begin{equation}\label{eq:tracevalue}
V(\state_0, \action_0, \state_1, \action_1,.., \state_N) =  \prod_{i=1..N-1} \prob_i \sum_{i=1..N-1} \reward(\state_i, \action_i)
\end{equation}

\noindent The optimal trace given a plan, i.e., the sequence of non-assumptive
actions from the trace, therefore has value:

\[
V^* = \max_{\prob_1, .., \prob_n} \prod_{i=1..N-1} \prob_i \sum_{i=1..N-1}
\reward(\state_i, \action_i),
\]

\noindent which is equal to the maximal 
additive contribution a trace of the plan can make to
Eq~\ref{eq:expectedvalue}.  In goal directed problems, where the only
non-zero reward is received at the first transition to a goal state,
the Eq~\ref{eq:tracevalue} criteria gives us the behaviour sought
by \system{FFR$_a$}~\cite{yoon:etal:2007}. Finally, it is
worth clarifying that although the above discussion has been in terms
of a trace of length $N$, there is nothing in our approach that
artificially limits the length of candidate sequential plans.

%% From the perspective of
%% gradient-based reinforcement learning systems, such as William's
%% REINFORCE and GPOMDP variants for planning~\cite{olivier:doug:2009},
%% under reasonable assumptions the optimal trace identifies a maximal
%% gradient step from a uniformly random soft-max policy.

\subsection{Decision-Theoretic Planning in Abstractions}

In a contingent session our switching planner solves a sensing problem
in an abstract processes defined in terms of the assumptive actions in
the current trace, and the $(\pp{:init})$ declaration that models the
underlying belief-state.

The focus for contingent planning is on sensing, we therefore give the
process in this session an augmented reward model.  This reflects the
value of performing sensing actions in the context of the trace
proposed by the preceding sequential deliberation. First, all rewards
from the original POMDP are retained. Then, for each $\assumptiveS{i}$
action scheduled by the current trace, we have a {\em dual}
$\assumptiveDT{i}$ so that $\forall\state\in\states$:

\[
\reward(\state, \assumptiveDT{i}) = \bigg\{ \begin{array}{ll}
\$(T_i) & \pp{if}~\;\;T_i \not\subseteq \state \\
\hat\$(T_i) & \pp{otherwise} \\
\end{array}
\]

\noindent where $\$(T_i)$ (resp. $\hat\$(T_i)$) is a positive
(negative) numeric quantity which captures the utility the agent
receives for correctly (incorrectly) rejecting an assumption. The
added dual actions are propositionally trivial -- i.e.,
$\forall\state\in\states$ $\poss(\assumptiveDT{i})
\subseteq \state$ and $\assumptiveDT{i}(\state) \equiv \state$.  
%%
In order to reduce the number of possible rejections we omit
assumptions that are not {\em active} with respect to the action,
\switchAction, whose scheduled execution switched the system into
the contingent session. An assumption $\assumptiveS{i}$ is active if
\switchAction\ is not applicable in the sequential model unless
$\assumptiveS{i}$ is scheduled in the trace prefix
to \switchAction. For example, if $\pp{DORA}$'s trace is:

\[
\begin{array}{l}
\actions^{\circ}(.8;(=(\pp{is-in}~\pp{box})\pp{office}));\\
\actions^{\circ}(.3;(=(\pp{is-in}~\pp{cup})\pp{kitchen}));\\
(\pp{look-for-object}~\pp{DORA}~\pp{box}~\pp{office});\\
(\pp{look-for-object}~\pp{DORA}~\pp{cup}~\pp{kitchen});\\
(\pp{report-is-in}~\pp{box}~\pp{office}); \\
(\pp{report-is-in}~\pp{cup}~\pp{kitchen})
\end{array}
\]

\noindent Taking the switching action \switchAction\ to be
$(\pp{look-for-object}~\pp{DORA}~\pp{box}~\pp{office})$, we have that
$\actions^{\circ}(.3;(=(\pp{is-in}~\pp{cup})\pp{kitchen}))$ is not
active, and therefore exclude
$\actions^{\bullet}(.3;(=(\pp{is-in}~\pp{cup})\pp{kitchen}))$ from the
POMDP posed to the contingent session.

We also include another propositionally trivial action
$\actions.\poss(\switchAction)$ with the reward property:

@@

For the purposes of successive switching, execution of either a dual
action of the form $\assumptiveDT{i}$, or the
$\actions.\poss(\switchAction)$, returns control to a sequential
session. If a dual is executed, then the sequential planner must
replan with $\bstate_0$ equal to the underlying
belief-state. Otherwise, if $\actions.\poss(\switchAction)$ is
executed, then the current sequential plan is executed until further
sensing is scheduled, or to completion. Lastly, we note that in our
project the rewards, i.e., functions
$\$,\hat\$:2^\propositions\to\Re$, for the POMDP posed to the
contingent session are sourced from a motivational component of our
robotic architecture. In this paper, for $\assumptiveDT{i}$ actions we
set $\$(x)$ to be a positive constant, and have $\hat\$(x)= - \$(x)(1
- \prob) / \prob$ where $\prob$ is the probability that $x$ is true. For
$\actions.\poss(\switchAction)$ actions we have $\hat\$(x)= -
\$(x)\prob/(1-\prob)$.\footnote{These values were chosen to encourage
  sensing.}

@@

% \small
% \begin{tabtt}
% (\=:init (= (is-in DORA) kitchen) \+ \\
%        (probabilistic \=.8 (= (is-in box) office)  \\
% 		      \>.2 (= (is-in box) kitchen))) \\
% \end{tabtt}
% \normalsize

% \noindent and therefore, the belief-state $\bstate_0$ is:

% \small
% \begin{tabular}{cccc}
% \hline
% Probability & (is-in DORA)  & (is-in box)  & (is-in cup) \\
% \hline
% .8 & kitchen & office & \# \\
% .2 & kitchen & kitchen & \# \\
% \hline
% \end{tabular}
% \normalsize

 % With
% regards to abstraction, the contingent and serial sessions apply the
% same applicability conditions ---i.e., that each effect and condition
% $\prop$ is specified, $\prop\in\state$--- to actions from the DTPDDL
% problem description.

>>>>

\subsection{Sequential Sessions}

We consider only deterministic-action POMDs, where all the uncertainty
in state and action is expressed in the $(\pp{:init})$ declaration. In
sequential sessions, this declaration is used by our approach to
define a starting state, and the set of state-assumptions available to
sequential planning.  Writing \#\ if the value of a proposition is
unspecified, continuing our example from the previous section, we have
the following assumptions:

\small
\begin{tabular}{cccc}
\hline
Probability & (is-in R2D2)  & (is-in box)  & (is-in cup) \\
\hline
%% .24 & kitchen & office & office \\
%% .06 & kitchen & kitchen & office \\
%% .56 & kitchen & office & kitchen \\
%% .14 & kitchen & kitchen & kitchen \\
.7 & kitchen & \# &  kitchen\\
.3 & kitchen & \# & office \\
.8 & kitchen & office & \# \\
.2 & kitchen & kitchen & \# \\
1.0 & kitchen & \# & \# \\
\hline
\end{tabular}
\normalsize

\noindent Each assumption corresponds to one distinct {\em
relaxed} visitation of the root term of $(\pp{:init})$. In this
relaxed case, a conjunctive term is visited iff its atomic subterms
are visited, and zero or one of its immediate probabilistic subterms
are visited. For a sequential session, the starting state corresponds
to the unique assumption with probability $1$ -- I.e., the union of
facts from the root conjunctive term. In our example, the starting
state is:

\small
\[
\begin{array}{l}
\state_0 \equiv \{(=(\pp{is-in}~\pp{R2D2})~\pp{kitchen}),\\
\;\;(=(\pp{is-in}~\pp{box})~\#), (=(\pp{is-in}~\pp{cup})~\#)\}.
\end{array}
\]
\normalsize

We also annotate DTPDDL model with an assumptive action
$\assumptiveS{i}$ for each element $\prob_i (T_i)$, of each
probabilistic term from $(\pp{:init})$. Here, $\assumptiveS{i}$ can be
executed if no $\assumptiveS{j}$, $j \neq i$, has been executed from
the same probabilistic term, and, either
$(\pp{probabilistic}~..\prob_i~(T_i)..)$ is in the root conjunct, or
it occurs in $T_k$ for some executed $\assumptiveS{k}$. 
%%
We add additional constraints that forbid the scheduling of
assumptions about facts after actions with preconditions, or effects,
that mention those facts. For example, the robot cannot assume it is
plugged into a power source immediately after it unplugs itself.
%%
Executing $\assumptiveS{i}$ in a state $\state$ effects a transition
to a successor state $\state^{T_i}$ with probability $\prob_i$, and
$\state^\bot$ with probability $1 - \prob_i$. Here, $\state^{T_i}$ is
the union of $\state$ with atomic terms from $T_i$. State
$\state^\bot$ is an added sink.

We now describe the optimisation criteria used during sequential
sessions. Where $\prob_i$ is the probability that the $i^{th}$
sequenced action $\action_i$ from trace,
$\state_0, \action_0,\state_1, \action_1,.., \state_N$, does not
transition to $\state^\bot$, we have that the optimal trace has value:

%% define the value of a trace
%% $\state_0, \action_0, \state_1, \action_1,.., \state_N$ from the
%% deterministic model to equal:

%% \small
%% \begin{equation}\label{eq:tracevalue}
%% V(\state_0, \action_0, \state_1, \action_1,.., \state_N) =  \prod_{i=1..N-1} \prob_i \sum_{i=1..N-1} \reward(\state_i, \action_i)
%% \end{equation}
%% \normalsize

%% \noindent The
%% optimal trace given a plan, i.e., the sequence of non-assumptive
%% actions from the trace, therefore has value:

\small
\[
V^* = \max_N \max_{\state_0, \action_0,.., \state_N} \prod_{i=1..N-1} \prob_i \sum_{i=1..N-1}
\reward(\state_i, \action_i),
\]
\normalsize

\noindent  Here,  $\reward(\state_i, \action_i)$ is the instantaneous
reward ---i.e., can be negative for action costs, or positive for goal
achievement--- received for executing action $\action_i$ in state
$\state_i$. Finally, it is worth clarifying that in practise we do not
artificially limit the length of traces that the sequential planner
can consider. Moreover, in our setting there is always a finite
optimal trace.

\subsection{DT Sessions}

The DT session plans in a {\em small} abstract processes defined
according to the assumptive actions in the proceeding trace, and the
$(\pp{:init})$ declaration that models the underlying
belief-state. Targeted sensing is encouraged by augmenting the reward
model to reflect the value of knowing the truth about assumptions in
the trace proposed by the preceding sequential deliberation. In
detail, all rewards from the original POMDP are
retained. Additionally, for each $\assumptiveS{i}$ action scheduled by
the current trace, we have a {\em dual} $\assumptiveDT{i}$ so that
forall states $\state$:

\small
\[
\reward(\state, \assumptiveDT{i}) = \bigg\{ \begin{array}{ll}
\$(T_i) & \pp{if}~\;\;T_i \not\subseteq \state \\
\hat\$(T_i) & \pp{otherwise} \\
\end{array}
\]
\normalsize

\noindent where $\$(T_i)$ (resp. $\hat\$(T_i)$) is a 
relatively\footnote{Compared to rewards from the original process
model.} small positive (negative) numeric quantity which captures the
utility the agent receives for correctly (incorrectly) rejecting an
assumption. Dual action execution is modelled as a self-transition --
I.e., the process remains in the state from which the dual is
executed.
%%
Furthermore, if \switchAction\ is the action that switched the system
to a DT session, then an assumption $\assumptiveS{i}$ is only {\em
active} if
\switchAction\ is not applicable in the sequential model unless
$\assumptiveS{i}$ is scheduled in the trace prefix
to \switchAction. We only include the duals of active assumptions in
DT sessions. For example, if the trace is:

\small
\[
\begin{array}{l}
\actions^{\circ}(.8;(=(\pp{is-in}~\pp{box})\pp{office}));\\
\actions^{\circ}(.3;(=(\pp{is-in}~\pp{cup})\pp{kitchen}));\\
(\pp{look}~\pp{box}~\pp{office});
(\pp{look}~\pp{cup}~\pp{kitchen});\\
(\pp{report}~\pp{box}~\pp{office}); 
(\pp{report}~\pp{cup}~\pp{kitchen})
\end{array}
\]
\normalsize

\noindent Taking the switching action \switchAction\ to be
$(\pp{look}~\pp{box}~\pp{office})$, we have that
$\actions^{\circ}(.3;(=(\pp{is-in}~\pp{cup})\pp{kitchen}))$ is not
active, and therefore exclude
$\actions^{\bullet}(.3;(=(\pp{is-in}~\pp{cup})\pp{kitchen}))$ from the
POMDP posed to the DT session.

Given \switchAction, we also include another self-transition action
$\actions.\poss(\switchAction)$ with the reward property:

@@

\noindent We note that in our project the dual rewards are sourced
from a motivational component of our robotic architecture. In this
paper, for $\assumptiveDT{i}$ actions we set $\$(x)$ to be a positive
constant, and have $\hat\$(x)= - \$(x)(1 - \prob) / \prob$ where
$\prob$ is the probability that $x$ is true. For
$\actions.\poss(\switchAction)$ actions we have $\hat\$(x)= -
\$(x)\prob/(1-\prob)$
%%
Finally, for the purposes of successive switching, execution of either a dual
action of the form $\assumptiveDT{i}$, or the
$\actions.\poss(\switchAction)$, returns control to a sequential
session, which then continues from the underlying belief-state. If an
assumption was rejected, we prohibit that sequential session from
making it again.
%%

@@
